---
title: Des Moines Data
author: Hunter Rose
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(choroplethr)
library(choroplethrMaps)
library(lubridate)
library(ggrepel)
library(osmdata)
library(sf)
library(ggmap)
```

```{r}
polk_crash_data = read.csv('Polk_County_Data.csv')
polk_vehicle_data = read.csv('Polk_County_Vehicle_Data.csv')
```

```{r}
polk_crash_data = polk_crash_data %>% 
    mutate(CRASH_DATE = as_datetime(as.character(CRASH_DATE)),
        LITERAL = as.character(LITERAL),
        DRUGALC = factor(DRUGALC, levels=rev(c('None Indicated', 'Refused', 
            'Alcohol (<Statutory)', 'Drug/Alcohol (< Statutory)', 
            'Drug', 
            'Alcohol (Statutory)', 'Drug/Alcohol (Statutory)', 
            'Under Influence of Alcohol/Drugs/Medications'))))
        

weather = c('Blowing sand, soil, dirt', 'Blowing Snow', 'Clear', 'Cloudy', 'Fog, smoke, smog',
            'Freezing rain/drizzle', 'Rain', 'Severe Winds', 'Sleet, hail', 'Snow')
levels(polk_crash_data$WEATHER)[!levels(polk_crash_data$WEATHER) %in% weather] = NA
surface = c('Dry', 'Gravel', 'Ice/frost', 'Mud, dirt', 'Oil', 'Sand', 'Slush', 'Snow', 'Wet')
levels(polk_vehicle_data$CSURFCOND)[!levels(polk_vehicle_data$CSURFCOND) %in% surface] = NA

polk_crash_data = polk_crash_data %>%
  filter(X > -94 & X < -93 & Y < 42 & Y > 41.3)
polk_vehicle_data = polk_vehicle_data %>%
  filter(X > -94 & X < -93 & Y < 42 & Y > 41.3)

```

```{r}
str(polk_vehicle_data)
str(polk_crash_data)
```


```{r}
polk_crash_data %>%
  ggplot(aes(WEATHER)) + geom_bar() + theme(axis.text.x = element_text(angle=60, hjust = 1))
polk_vehicle_data %>%
  ggplot(aes(CSURFCOND)) + geom_bar() + theme(axis.text.x = element_text(angle=60, hjust = 1))
```

After looking at the Weather graph, we can see that around 90,000 of crashes occur during clear/cloudy conditions. This was somewhat surprising to see such high occurences with conditions that aren't difficult to drive in. Rain and snow both combine for around 17,000 crashes, a small comparison to clear/cloudy. When looking at surface conditions, we can see that most crashes are on dry concrete. We see wet conditions being the next largest category, with snow and ice/frost making up the third spot. I thought that we would see more crashes with bad conditions, but it could be because people tend to stay off the roads when there is bad weather.


```{r}
polk_crash_data %>%
  ggplot(aes(DRUGALC)) + geom_bar() + theme(axis.text.x = element_text(angle=60, hjust = 1))

levels(polk_crash_data$DRUGALC)
drug_alc = c('Under Influence of Alcohol/Drugs/Medications', 'Drug/Alcohol (Statutory)', 'Alcohol (Statutory)',
             'Drug', 'Drug/Alcohol (< Statutory)', 'Alcohol (<Statutory)', 'Refused')
levels(polk_crash_data$DRUGALC)[!levels(polk_crash_data$DRUGALC) %in% drug_alc] = NA

polk_crash_data %>%
  select(DRUGALC) %>%
  drop_na(DRUGALC) %>%
  ggplot(aes(DRUGALC)) + geom_bar() + theme(axis.text.x = element_text(angle=60, hjust = 1))
```

At first glance into the drugs/alcohol category, we can see that the majority of the crashes indicated there was no substances involved. After removing this level, we can get a better look at those driving under the influence. Once filtered, the majority of crashes that involved substances was alcohol. An interesting category to look at in this graph is those that refused to take a test. Depending on the crash evidence, refusing to take a breathalyzer can get you out of a DUI. It can also get you into more trouble and long-term issues such as suspended license or jail time. So to refuse a test is sort of a gamble when it comes to the law.


```{r}
polk_vehicle_data %>%
  group_by(SPEEDLIMIT) %>%
  ggplot(aes(x=X, y=Y, color=SPEEDLIMIT)) + geom_point()
```

When looking at this plot, we can start to see where the highways are and where the more residential areas are.

```{r}
des_map <- get_map(getbb("Des Moines"))

fatal <- polk_vehicle_data %>%
  filter(FATALITIES > 0) %>%
  group_by(FATALITIES)


ggmap(des_map) + geom_point(data = fatal, aes(x=X, y=Y, color=FATALITIES))
```

When there are fatalities in a crash, we can see that most of them are only one person. There are some places that have some type of consistency for crashes that lead to fatalities, those around the interstate on/off ramp, along with those around 2nd Ave. 

```{r}
polk_vehicle_data %>%
  group_by(DRIVERAGE) %>%
  filter(DRIVERAGE < 95) %>%
  ggplot(aes(x=DRIVERAGE, color=DRIVERGEN)) + geom_histogram() + geom_vline(aes(xintercept=mean(DRIVERAGE)), color = "blue", linetype="dashed", size=1)
```

This graph has shown that the average age of the person causing the crash is around 38 years. We also see a very high amount of crash in the 16-23 years range, with a steady decrease in crashes per age as age increases outside of around the 43 mark. As far as gender goes, we can see that there is a slight majority of males than females. 

```{r}
streets = polk_crash_data %>% mutate(total=X*X+Y*Y) 

scnt = streets %>% 
    group_by(total) %>% 
    summarize(count = n())
    
streets = streets %>% 
    left_join(scnt, by='total') %>% 
    arrange(desc(count))

str(streets %>% distinct(LITERAL, .keep_all=TRUE))


topstreets = streets %>% 
    distinct(total, .keep_all=TRUE) %>% 
    distinct(LITERAL, .keep_all=TRUE) %>% 
    filter(LITERAL %in% head(LITERAL, 10))

des_map <- get_map(getbb("Des Moines"))

ggmap(des_map) + geom_point(data = topstreets, aes(x=X, y=Y, color=count)) + 
  geom_text(data = topstreets, aes(x=X, y=Y, label = LITERAL), hjust=1, vjust=0, size=3, angle=20)
```

These are the most dangerous intersections in the polk county area. The street with the most crashes is US 69/SE 14th ST & E Park Ave. 

